---
title: "An Overview of a Global Pandemic"
author: 'Group 4: Almost the Data Sisters'
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    source: embed
runtime: shiny
    
---
# Overview

```{r setup, include=FALSE}
library(flexdashboard)
library(shiny)
library(dplyr)
library(plotly)
library(leaflet)

# Upload all data for all countries
covid_df <- readRDS(file="country.info.RDS")

# Summarize data to get mortality and morbidty for an entire region
by_region <- covid_df %>% 
  group_by(region) %>%
  filter_at(vars(confirmed,deaths,population), all_vars(!is.na(.))) %>%
  summarise(regional_mortality = ((sum(deaths)/sum(population))*100000), regional_morbidity = ((sum(confirmed)/sum(population))*100000))
```

Inputs {.sidebar}
-----------------------------------------------------------------------
### Overview 

**COVID-19** is an infectious disease caused by the SARS-CoV-2 virus. The emergence of the COVID-19 global pandemic has received much attention in the scientific world, prompting researchers to actively explore the potential socioeconomic factors affecting the spread of COVID-19 and and mortality rates. 

This dashboard is intended to provide a visual representation of selected socioeconomic 
factors that may impact the spread of the disease as well as mortality rates across countries. 

<br>
**How to Use:** The interactive map on the right provides country-specific status relating to COVID-19. You may click to drag the map around, or click the '+' and '-' buttons to zoom in and out. You may hover over countries to reveal a popup with information about the specific country. 

Row {data-height=150}
-----------------------------------------------------------------------

### Value: Total Cases

```{r valuebox_total}
total_confirmed <- sum(covid_df$confirmed, na.rm=TRUE)
flexdashboard::valueBox(formatC(total_confirmed, format = "d", big.mark = ","), 
                        caption = 'Total Confirmed Cases to Date',
                        icon = "fas fa-capsules", 
                        color = "#fcbe42")
```


### Value: Total Deaths
```{r valuebox_deaths}
total_deaths <- sum(covid_df$deaths, na.rm=TRUE)
flexdashboard::valueBox(formatC(total_deaths, format = "d", big.mark = ","),
                        caption = 'Total Deaths',
                        icon = "fas fa-dizzy",
                        color = "#86dff7")
```


### Value: Reported Recoveries
```{r valuebox_recovery}
total_recovered <- sum(covid_df$recovered, na.rm=TRUE)
flexdashboard::valueBox(formatC(total_recovered, format = "d", big.mark = ","),
                        caption = 'Reported Recoveries',
                        icon = "fas fa-heart",
                        color = "#95d468")
```

Row {.tabset}
-----------------------------------------------------------------------

### Global Summary: Confirmed Cases

```{r spdf_data}
library(leaflet)
library(rgdal)
library(dplyr)
library(RColorBrewer)

# Download world shape file 
download.file("http://thematicmapping.org/downloads/TM_WORLD_BORDERS_SIMPL-0.3.zip", destfile=paste0(getwd(), "/world_shape_file.zip"))


# Read shape file with rgdal into a spdf (spatial polygon dataframe) object 
world_spdf <- readOGR( 
  dsn= paste0(getwd(),"/world_shape_file/") , 
  layer="TM_WORLD_BORDERS_SIMPL-0.3",
  verbose=FALSE)
```


```{r combine_data}
# Clean & prepare covid_df for merging
covid_cleaned <- covid_df %>%
  rename("ISO3" = "iso3c") %>%
  # Select columns of interest
  select(ISO3, confirmed, deaths, recovered, morbidity_rate, mortality_rate, recovery_rate)

# Combine spdf & covid_df datasets based on ISO3 code
world_spdf_join <- merge(world_spdf, covid_cleaned, by="ISO3")

# Clean the data object
world_spdf_join@data$confirmed <- as.numeric(as.character(world_spdf_join@data$confirmed)) / 1000000 %>% round(2)
world_spdf_join@data$deaths <- as.numeric(as.character(world_spdf_join@data$deaths)) / 1000 %>% round(2)
```


```{r map_confirmed}
# Create a color palette with handmade bins
mybins <- c(0, 1, 5, 10, 20, 50, Inf)
mypalette <- colorBin(palette="YlOrBr",
                      domain=world_spdf_join@data$confirmed,
                      na.color="transparent",
                      bins=mybins)

# Prepare text for popup
mytext <- paste(
    "Country: ", world_spdf_join@data$NAME,"<br/>", 
    "Confirmed Cases (M): ", round(world_spdf_join@data$confirmed, 2), "<br/>", 
    "Morbidity Rate: ", world_spdf_join@data$morbidity_rate, 
    sep="") %>%
  lapply(htmltools::HTML)

# Plot choropleth map with leaflet
m <- leaflet(world_spdf_join) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng=0, lat=30, zoom=2) %>%
  addPolygons(fillColor = ~mypalette(confirmed), 
              stroke = FALSE,
              fillOpacity = 0.5,
              smoothFactor = 0.5,
              weight = 0.3,
              color = "white",
              label = mytext,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "13px",
                direction = "auto"
              )
  ) %>%
  
  # Add legend to the map
  addLegend(pal = mypalette, 
            values = ~confirmed, 
            opacity=0.9, 
            title = "COVID-19 Confirmed Cases (M)", 
            position = "bottomleft")

m
```

### Deaths

```{r map_deaths}
# Create a color palette with handmade bins
mybins <- c(0, 1, 10, 30, 50, 100, 500, Inf)
mypalette <- colorBin(palette="YlGnBu",
                      domain=world_spdf_join@data$deaths,
                      na.color="transparent",
                      bins=mybins)

# Prepare text for popup
mytext <- paste(
    "Country: ", world_spdf_join@data$NAME,"<br/>", 
    "Deaths (K): ", round(world_spdf_join@data$deaths, 2), "<br/>", 
    "Mortality Rate: ", world_spdf_join@data$mortality_rate, 
    sep="") %>%
  lapply(htmltools::HTML)

# Plot choropleth map with leaflet
m <- leaflet(world_spdf_join) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng=0, lat=30, zoom=2) %>%
  addPolygons(fillColor = ~mypalette(deaths), 
              stroke = FALSE,
              fillOpacity = 0.5,
              smoothFactor = 0.5,
              weight = 0.3,
              color = "white",
              label = mytext,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "13px",
                direction = "auto"
              )
  ) %>%
  
  # Add legend to the map
  addLegend(pal = mypalette, 
            values = ~deaths, 
            opacity=0.9, 
            title = "COVID-19 Deaths (K)", 
            position = "bottomleft" )

m
```

# Regional Data

Inputs {.sidebar}
-----------------------------------------------------------------------

On this page, the graphs represent the cases (morbidity rate) and deaths (mortality rate) per 100,000 individuals in both regions defined by the World Bank and country economies that rest within those regions.

Here we can see large differences in rates between regions around the world and countries within the same region. Factors that may explain these differences include:

* **Limited testing capacities:** Some nations lack the resources required for testing and personnel, leading to under reporting of case numbers. Death rates due to COVID-19 in comparison to overall mortality rates have been documented to be a more accurate method for determining the true presence of COVID-19 ([Pulmonology](https://www.sciencedirect.com/science/article/pii/S253104372030129X))

* **Population Density:** Although some regions may share similar populations, those with greater population density may be subject to increased transmission. Higher case numbers and rates can be expected in regions with greater density. ([PLOS ONE](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0242398))

```{r region}
# region selection
selectInput("region", label = h3("Region of Interest:"),
            choices = c("East Asia & Pacific",
                        "Europe & Central Asia",
                        "North America", 
                        "Latin America & Caribbean",
                        "Middle East & North Africa",
                        "South Asia",
                        "Sub-Saharan Africa"), selected = "All")

```

**How to Use:** To view the morbidity and mortality rates for countries within a specific region, choose a region from the drop down menu. For comparing countries with similar rates in Charts 3 and 4, you can click and drag to select a portion of them.

Row {.tabset}
---

### **Chart 1:** COVID-19 cases per 100,000 individuals across all regions

```{r morbidity_region}

# static graph for morbidity across all regions
plot_ly(by_region, 
        x = ~region, 
        y = ~regional_morbidity, 
        type = "bar", 
        text = ~regional_morbidity,
        textposition = "outside", 
        textformat = "digits",
        texttemplate = "%{y:d}",
        textcolor = "black",
        hovertext = ~ paste0(region, "<br>", regional_morbidity),
        hoverinfo = "text",
        color = ~region,
        colors = "#fec44f") %>% 
  layout(yaxis = list(title = "<b>Cases per 100,000 Individuals", type = "log", fixedrange = T), 
         xaxis = list(title = "<b>World Bank Region", tickangle = 0, tickformat = "digits", tickfont = list(size = 11.3), fixedrange = T),
         uniformtext=list(minisize=8, mode = "hide"),
         showlegend = F)
```

### **Chart 2:** COVID-19 deaths per 100,000 individuals across all regions

```{r mortality_region}

# static graph for mortality across all regions
plot_ly(by_region, 
        x = ~region, 
        y = ~regional_mortality, 
        type = "bar", 
        text = ~regional_mortality,
        textposition = "outside", 
        texttemplate = "%{y:d}",
        hovertext = ~ paste0(region, "<br>", regional_mortality),
        hoverinfo = "text",
        color = ~region,
        colors = "#225ea8") %>% 
  layout(yaxis = list(title = "<b>Deaths per 100,000 Individuals</b>", type = "log", fixedrange = T), 
         xaxis = list(title = "<b>World Bank Region</b>", tickangle = 0, tickfont = list(size = 11.3)),
         uniformtext=list(minisize=8, mode = "hide"),
         showlegend = F)

```

Row
---

### **Chart 3:** COVID-19 cases per 100,000 individuals by country in a given region
```{r morbidity_plot}

# make dataset reactive based on region choice
morb_data <- reactive({
  covid_df %>%
    filter(region == input$region)
})

# reactive graph for country morbidity within individual regions 
renderPlotly({
  plot_ly(morb_data(),
          x = ~country,
          y = ~morbidity_rate,
          type = "bar",
          hovertext = ~ paste0(country, "<br>", morbidity_rate),
          hoverinfo = "text",
          color = ~country,
          colors = "#fec44f" ) %>%
    layout(yaxis = list(title = "<b>Cases per 100,000 Individuals</b>", fixedrange = T),
           xaxis = list(title = "<b>Country Economy</b>", tickangle = 45, vjust = 1,  categoryorder = "total ascending"),
           uniformtext=list(minisize=8, mode = "hide"),
           showlegend = F)
})

```

### **Chart 4:** COVID-19 deaths per 100,000 individuals by country in a given region
```{r mortality_plot}

# make dataset reactive based on region choice
mort_data <- reactive({
  covid_df %>%
    filter(region == input$region)
})

# reactive graph for country mortality within individual regions 
renderPlotly({
  plot_ly(mort_data(),
          x = ~country,
          y = ~mortality_rate,
          type = "bar",
          hovertext = ~ paste0(country, "<br>", mortality_rate),
          hoverinfo = "text",
          color = ~country,
          colors = "#225ea8" )%>%
    layout(yaxis = list(title = "<b>Deaths per 100,000 Individuals</b>", fixedrange = T),
           xaxis = list(title = "<b>Country Economy</b>", tickangle = 45, categoryorder = "total ascending"),
           uniformtext=list(minisize=8, mode = "hide"),
           showlegend = F)
})


```


# Socioeconomic Data

Inputs {.sidebar}
-----------------------------------------------------------------------


# About

Row {data-height=500}
-----------------------------------------------------------------------
### Dashboard

**An Overview of a Global Pandemic Dashboard**  

This dashboard provides an overview of the worldwide pandemic of coronavirus
disease 2019 (COVID-19), with an emphasis on the geographic locations and 
socioeconomic status of different countries, and how these factors may impact 
the spread of the disease across the globe. The dashboard was constructed in R with [Rmarkdown](https://rmarkdown.rstudio.com/) and the [flexdashboard](https://pkgs.rstudio.com/flexdashboard/) framework.

<br>
**Data Source** 

Data is retrieved from the [World Bank](https://datahelpdesk.worldbank.org/) and the [Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19) Coronavirus repository.  

The **World Bank** is an international financial organization that endeavors to reduce extreme poverty and promote sustainable prosperity by providing loans and grants to low and middle-income countries. World Bank provides free and open access to global development data, which are collected and reported annually or less often, depending on the specific indicator. 

The COVID-19 Data Repository operated by the **Johns Hopkins University Center for Systems Science and Engineering (JHU CCSE)** collects information from aggregated data sources like the World Health Organization (WHO) and the European Centre for Disease Prevention and Control (ECDC). The data on confirmed COVID-19 infection cases and deaths are updated on a daily basis, while the data on recoveries is no longer updated as of August 4th, 2021. For the construction of this dashboard, we accessed COVID-19 information of various countries from 2020 including total cases, deaths, and recoveries.

<br>
**Assumptions**  

Not all socioeconomic indicators contained up-to-date data from 2020. In these cases, data from earlier years were used for our analysis and the construction of the dashboard. Therefore, it was assumed that the values of these indicators have not changed significantly between their year of measurement and 2021. 

<br>
**Packages** 

The following packages were used to construct the dashboard:

* Data retrieval: httr, jsonlite, wbstats
* Data manipulation: countrycode, dplyr, tidyr, tidyverse
* Dashboard interface: flexdashboard, shiny 
* Visualization: ggplot, ggcorrplot, ggpubr, ggthemes, leaflet, plotly, multcompView, corrplot, cowplot, plotrix


Row {data-height=700}
-----------------------------------------------------------------------
### Dataset
    
```{r}
DT::datatable(
  covid_df,
  filter = "top",
  extensions = "Scroller",
  rownames = FALSE,
  style = "bootstrap",
  class = "compact",
  width = "100%",
  options = list(deferRender = TRUE, scrollY = 300, scroller = TRUE))
```


# Citation

Lau, H., et al. “Evaluating the Massive Underreporting and Undertesting of COVID-19 Cases in Multiple Global Epicenters.” Pulmonology, no. 2, Elsevier BV, Mar. 2021, pp. 110–15. Crossref, doi:10.1016/j.pulmoe.2020.05.015.

Wong, David W. S., and Yun Li. “Spreading of COVID-19: Density Matters.” PLOS ONE, edited by Bing Xue, no. 12, Public Library of Science (PLoS), Dec. 2020, p. e0242398. Crossref, doi:10.1371/journal.pone.0242398.





